### 2018-03-27 submission/manuscript.Rmd
final intro fixes and first discussion fix
update discussion transitions except last paragraph

In this study we show that the number of OTUs, error rate, and chimera prevalence are HiFi DNA polymerase dependent [Figure 1-2 & 4-5]. These differences are important because many diversity metrics rely on the number of OTUs or other measures dependent on error rate and chimera prevalence as part of their metric calculations (e.g. richness). Importantly, many of these differences are due to undetected chimeras that cannot be fully removed using standard bioinformatic approaches. This suggests that some of the diversity differences between studies can be attributed to differences in HiFi DNA polyermase used. Interestingly, the earlier detection of differences in total number of OTUs between HiFi DNA polymerases in the mock versus fecal samples might indicate that high biomass samples may underestimate the biases present within low biomass samples.

Although the variation in error rate and chimera prevalence may be due to the DNA polymerase family, the highest and lowest chimera rates both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014]. Additionally, based on the material safety data sheet (MSDS) the differences between the two HiFi DNA polymerases are not immediately apparent. Both HiFi DNA polymerases contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody. With everything else being equal, it is possible that differences in how the recombinant *Taq* was generated could be a contributing factor for the differences observed between the HiFi DNA polymerases.

There were few differences in bacterial community composition based on HiFi DNA polymerase. The data also suggests that there were no differences in the overall bacterial community composition for sub-sampling depth or cycle number used. One possible reason for this outcome was that our study did not have enough power to detect differences due to low sample number in each group. Another reason was that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed HiFi DNA polymerase dependent differences in chimera prevalence. Metrics that simply  look at presence/absence of OTUs (e.g. Jaccard [@Real1996]) may be less robust to chimera prevalence and by extension total number of OTU differences in HiFi DNA Polymerases. When choosing a distance metric careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account.

Similar to using different preservation methods or different DNA extraction kits, the type of HiFi DNA polymerase can add bias to the observed bacterial community. The sequence error introduced by the HiFi DNA polymerase is small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014]. However, the chimarea prevalence for some HiFi DNA polymerases (e.g. Platinum) are relatively large and might be greater than the oberseved biological variation within a specific study. Within the larger context of the different 16S rRNA gene sequencing steps, the choice of HiFi DNA polymerase can be as important a consideration as either preservation or DNA extraction method used.

>>>>

In this study we show that the number of OTUs, error rate, and chimera prevalence depends on HiFi DNA polymerase [Figure 1-2 & 4-5]. These differences are important because many diversity metrics rely on the number of OTUs or other measures dependent on error rate and chimera prevalence as part of their metric calculations (e.g. richness). Additionally, the earlier detection of differences in total number of OTUs between HiFi DNA polymerases in the mock versus fecal samples might indicate that high biomass samples may underestimate the biases present within low biomass samples. We observed that undetected chimeras that were not identified and removed using standard bioinformatic approaches cause many of these differences. This suggests that specific diversity differences between studies can be attributed to differences in HiFi DNA polyermase used. Based on our observations metrics that measure within sample diversity depend on HiFi DNA polymerase but this may not be the case for metrics that assess between sample diversity.

There were few differences that depend on HiFi DNA polymerase for between sample diversity, as measured by the Bray-Curtis index. Using this metric our observations found no differences in the overall bacterial community composition for sub-sampling depth or cycle number used. One possible reason for this outcome was that our study did not have enough power to detect differences due to low sample number in each group. Another reason was that many of the OTUs are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight chimeric OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed HiFi DNA polymerase dependent differences in chimera prevalence. Metrics that solely use presence/absence of OTUs (e.g. Jaccard [@Real1996]) may be less robust to chimera prevalence and by extension total number of OTU differences in HiFi DNA Polymerases. When choosing a distance metric careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account. With differences in the number of OTUs and chimera prevalence depending on HiFi DNA polymerase used, it might be easier to avoid specific DNA polymerase families altogether.

Although the variation in error rate and chimera prevalence may be due to the DNA polymerase family, the highest and lowest chimera rates both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014]. Additionally, based on the material safety data sheet (MSDS) the differences between the two HiFi DNA polymerases are not immediately apparent. Both HiFi DNA polymerases contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody. With everything else being equal, it is possible that differences in how the recombinant *Taq* was generated could be a contributing factor for the differences observed between the HiFi DNA polymerases. We are unlikely to avoid adding HiFi dependent bias to 16S rRNA gene sequencing results, however, these differences may not be large enough to mask the actual biological signal.

The majority of HiFi DNA polymerases we studied add a small increase in the number of OTUs and chimera prevalence that can be masked by biological differences. The sequence error introduced by the HiFi DNA polymerase is small and likely to be smaller than the biological variation within a specific study, which would be consistent with previous findings for preservation and DNA extraction methods [@preservation_Luo_2016; @preservation_Song_2016; @contamination_Salter2014]. However, the chimarea prevalence for some HiFi DNA polymerases (e.g. Platinum) are relatively large and might be greater than the oberseved biological variation within a specific study. The choice of HiFi DNA polymerase can be as important a consideration as either preservation or DNA extraction method used because similar to using different preservation methods or different DNA extraction kits, the type of HiFi DNA polymerase can add bias to the observed bacterial community. Although avoiding HiFi DNA polymerases that yield high number of OTUs and chimera prevalence might provide better standardization of results it does come at a cost.
